Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-12 _tex/paper2.tex
Built site for gh-pages

Thurstone introduced the Law of Comparative Judgment to measure
psychological traits of stimuli through pairwise comparisons
\citep{Thurstone_1927a, Thurstone_1927b}. In its general form, the
theory models single-judge comparisons across multiple, potentially
correlated stimuli. Each comparison produces a dichotomous outcome
indicating which stimulus the judge perceives as having a higher trait
level. However, Thurstone identified one key issue in this model: it
required estimating more ``unknown'' parameters than the number of
available pairwise comparisons \citep{Thurstone_1927b}. To address this
issue and facilitate the theory's practical applicability, he formulated
five cases derived from the general form, each progressively
incorporating several simplifying assumptions.

@@

Furthermore, because Thurstone's primary goal was to produce a ``coarse
scaling'' of traits and to ``allocate the compared stimuli on this
continuum'' \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
how to use the resulting trait estimates for statistical inference or
hypothesis testing. The CJ tradition has attempted to address this gap
by separating trait estimation from hypothesis testing. Specifically, CJ
studies often rely on point estimates of the traits (i.e., the BTL
scores)--or their transformations--to conduct hypothesis tests. However,
the statistical literature cautions against this practice, as it can
introduce bias into the analysis and compromise the reliability of the
resulting conclusions
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

To address these issues, this study extends Thurstone's general form
using a systematic, integrated approach that combines Causal and
Bayesian inference methods. It begins by formulating a conceptual model
represented by an SCM and a DAG. This model integrates Thurstone's core
theoretical principles--such as the discriminal processes (i.e., traits)
of stimuli--alongside key CJ experimental design features, including
judges' bias, sampling procedures, and comparison mechanisms. The study
then translates the SCM into a bespoke statistical model that addresses
key limitations of the BTL model. In particular, it enables the analysis
of CJ data when the assumptions of equal dispersion and zero correlation
do not hold (see Section~\ref{sec-theory-issue1a} and
Section~\ref{sec-theory-issue1b}), and when researchers seek to conduct
statistical inference or hypothesis testing (refer to
Section~\ref{sec-theory-issue2}). These methodological innovations
potentially enhance the reliability and validity of trait measurement
\citep{Perron_et_al_2015} and improve the accuracy of statistical
conclusions.

Beyond these potential benefits, the approach offers two key advantages.
First, it clarifies the interactions among all actors and processes
involved in CJ experiments. Second, it shifts the analytic paradigm from
passively accepting the assumptions of the BTL model to actively testing
their fit with observed data. Together, these advantages create a
principled framework for evaluating best practices in CJ experimental
designs--one that better aligns with the demands of contemporary CJ
assessment contexts \citep{Kelly_et_al_2022} and opens new avenues for
research.

@@

Second, although initial simulation tests of the statistical model have
yielded promising results (not shown), these findings may not generalize
across research contexts. We have argued that the model's ability to
estimate traits accurately and support valid statistical inferences
depends heavily on the specific research context--particularly on
whether the data and inferential goals align with the model's underlying
assumptions. That said, our approach provides tools for researchers to
test and evaluate these context-dependent CJ models directly.

@@

Fourth and final, researchers must also be proficient in Probabilistic
Programming Languages (PPLs) and Bayesian inference methods. This
expertise is essential for translating context-specific conceptual
models into bespoke statistical models that meaningfully address the
research questions at hand.

>>>>

Thurstone introduced the Law of Comparative Judgment to measure
psychological traits of stimuli through pairwise comparisons
\citep{Thurstone_1927a, Thurstone_1927b}. In its general form, the
theory models single-judge comparisons across multiple, potentially
correlated stimuli. Each comparison produces a dichotomous outcome
indicating which stimulus the judge perceives as having a higher trait
level. However, Thurstone identified one key challenge in this model: it
required estimating more ``unknown'' parameters than the number of
available pairwise comparisons \citep{Thurstone_1927b}. To address this
issue and facilitate the theory's practical applicability, he formulated
five cases derived from the general form, each progressively
incorporating several simplifying assumptions.

@@

Furthermore, because Thurstone's primary goal was to produce a ``coarse
scaling'' of traits and to ``allocate the compared stimuli on this
continuum'' \citep[pp.~269]{Thurstone_1927a}, his theory did not specify
how to use the resulting trait estimates for statistical inference. The
CJ tradition has attempted to address this gap by separating trait
estimation from the hypothesis testing process. Specifically, CJ studies
often rely on point estimates of the traits (i.e., the BTL scores)--or
their transformations--to conduct statistical inference. Although
separating these processes simplifies the analysis of CJ data, the
statistical literature cautions against this practice, as it can
introduce bias into the analysis and compromise the reliability of the
resulting conclusions
\citep{McElreath_2020, Kline_et_al_2023, Hoyle_et_al_2023}.

Thus, to address the limitations of Thurstone's Case V and the BTL
model--particularly their strong assumptions and the disconnect between
trait measurement and hypothesis testing--this study extends Thurstone's
general form using a systematic, integrated approach that combines
Causal and Bayesian inference methods. It begins by formulating a
conceptual model represented by an SCM and a DAG. This model integrates
Thurstone's core theoretical principles--such as the discriminal
processes (i.e., traits) of stimuli--alongside key CJ experimental
design features, including judges' bias, sampling procedures, and
comparison mechanisms. The study then translates the SCM into a bespoke
statistical model that addresses key limitations of the BTL model. In
particular, it enables the analysis of CJ data when the assumptions of
equal dispersion and zero correlation do not hold (see
Section~\ref{sec-theory-issue1a} and Section~\ref{sec-theory-issue1b}),
and when researchers seek to conduct statistical inference or hypothesis
testing (refer to Section~\ref{sec-theory-issue2}). These methodological
innovations potentially enhance the reliability and validity of trait
measurement \citep{Perron_et_al_2015} and improve the accuracy of
statistical conclusions.

Beyond these potential benefits, the approach offers two additional
advantages. First, it clarifies the interactions among all actors and
processes involved in CJ experiments. Second, it shifts the analytic
paradigm from passively accepting the assumptions of the BTL model to
actively testing their fit with observed data. Together, these
advantages create a principled framework for evaluating best practices
in CJ experimental designs--one that better aligns with the demands of
contemporary CJ assessment contexts \citep{Kelly_et_al_2022} and opens
new avenues for research.

@@

Second, although initial simulation tests of the statistical model have
yielded promising results (not shown), these findings may not generalize
across research contexts. We have argued that the model's ability to
accurately estimate traits and support valid statistical inferences
depends heavily on the specific research context--particularly on
whether the data and inferential goals align with the model's underlying
assumptions. That said, our approach provides tools for researchers to
test and evaluate these context-dependent CJ models directly.

@@

Fourth and final, researchers must also be proficient in Probabilistic
Programming Languages (PPLs) and Bayesian inference methods. This
expertise is essential for translating context-specific conceptual
models into bespoke statistical models that meaningfully address the
research questions at hand. While such translation is not exclusive to
Bayesian methods--frequentist approaches can also be used--we argue that
Bayesian methods offer a more accessible pathway, as they reduce the
need for additional expertise in optimization theory and techniques.
